---
title: "starmie: Basic Usage"
author: "Stuart Lee, Gerry Tonkin-Hill"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    keep_md: true
vignette: >
  %\VignetteIndexEntry{starmie Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Starmie: making population structure analyses easier 

A very common part of modern population genetics analysis is inferring
underlying population structure from genetic markers such as single
nucleotide polymorphisms (SNPs) or microsatellites. The two main methods
for this task are the Bayesian STRUCTURE algorithm or the frequentist
ADMIXTURE. We have found that processing the output of these programs and
performing meaningful inference and visualisation of the results is far more
difficult than it should be. This is why we wrote starmie. 

Some key features:

  * Parse output from STRUCTURE or ADMXITURE into a useable data structure in R.
  * Plot model diagnostics to perform inference on choice of 'K'.
  * Create the 'standard' STRUCTURE bar plot showing cluster memberships.
  * Plot model checks such as MCMC chains on the admixture coeffecient $\alpha$
    and the log-likelihood.
  * Interactively visualising output from STRUCTURE analysis

This vignette outlines how to use starmie to do basic tasks after running 
STRUCTURE at the command line.

## A basic STRUCTURE pipeline in R.

To use all the options in starmie for STRUCTURE output, we require 
for each run the 'out_f' file produced by the program and the logging information
so we can produce MCMC diagnostics. To get the latter the output of STRUCTURE
must be redirected to a file. Below we present an example of running STRUCTURE
in multiple runs for each K in parallel. We assume that the `mainparams` and
`extraparams`  files are in the same directory and that the user has access
to the STRUCTURE binary. Also make sure RANDOMIZE option is turned off,
so independent seeds can be set in each run.

```{r parallel-structure, eval=FALSE}
# use parallel library
library(parallel)
n.cores <- detectCores() # run on all possible cores machine has
# prelim setup
structure_binary <- "/path/to/structure"
# number of Ks to try
tryK <- 1L:20L
# number of runs for each K
n.runs <- 1L:20L

# output files (will be appended by out_f from structure)
out_files <- outer(n.runs, tryK, 
                   function(x,y) paste0("str_run_", 
                                        stringr::str_pad(x, width = 2, pad = 0), 
                                        "_K_", 
                                        stringr::str_pad(y, width = 2, pad = 0), 
                                        ".out")
                   )

log_files <- gsub("out", "log", out_files)

# create function to run structure, assumes that mainparams/extraparams 
# file is in same path as current wd in Rscript
run_structure <- function(out_file, log_file) {
  k <- as.integer(stringr::str_extract(out_file, "[0-9]{2}\\b"))
  return(system(paste(structure_binary, 
                      "-K", k, 
                      "-D", round(runif(1) * 1e8), 
                      "-o", out_file, "&>", log_file)))
}

# prepare run with mcapply
mcmapply(run_structure, out_files, log_files, 
                    mc.cores = n.cores, mc.set.seed = TRUE)

```


## Parsing STRUCTURE files, the 'struct' object

The basic unit of analysis for `starmie` is the `struct` object, which contains
the model information in the STRUCTURE out_f file and optionally the logging
information for the MCMC diagnostics. As an example, we have run STRUCTURE
on [simulated microsatellite data](http://pritchardlab.stanford.edu/software/example-data/locprior.str) 
from the STRUCTURE example data and save the out_f. To create a `struct` object
from a run we use the following:
```{r struct-example}
library(starmie)
# path to file name
k6_file <- system.file("extdata/microsat_testfiles/", "locprior_K6.out_f", 
                       package = "starmie")
# create struct object
k6_msat <- loadStructure(k6_file)

k6_msat
str(k6_msat)
```


data structure for the `starmie` package is the 

## Diagnostics for 'struct' objects

## Inference on K is hard

## Why you're here: the bar plot.


## Why starmie?
There's a reason.

# session info
```{r session-info}
sessionInfo()
```
